The kinematics of the Local Group in a cosmological context

### 2013-02-11 paper/ms.tex
updated version

\begin{abstract}
Recent observations constrained the tangential velocity of M31 with
respect to the Milky Way (MW) to be $v_{\rm M31,tan}<34.4$ \kms and
the radial velocity to be in the range $v_{\rm M31,rad}=-109\pm
4.4$\kms \citep{vanderMarel12}. In this study we use a large volume
high resolution N-body cosmologicalsimulation (BOLSHOI) to
statistically study this kinematics in the context of the $\Lambda$CDM
cosmology and three constrained simulations designed to reproduce the
Local Group (LG) and the local cosmic neighborhood. A LG is defined
here as  composed by two dominant separate halos hosting the MW and
the M31 galaxy fulfilling certain isolation criteria. An ensemble  of
all BOLSHOI pairs fulfilling this selection is constructed, and also a
subsample of it which obeys  more stringent observational constraints
on the mass and separation of of the MW and M31 (hereafter the full-
and sub- samples). The comparison of the ensembles of simulated pairs
with the observed LG has been done with respect to the observed radial
and tangential,  the reduced orbital energy ($e$) and angular momentum
($l$) of the LG, and with respect to the dimensionless spin parameter,
$\lambda$. Our main results are: 
%
%
% Jaime here I suggest to give in a very brief way the numbers of Table 2 that I have suggested.
%
The three constrained LGs are more consistent with the unconstrained
sample of LGs rather then with the observed LG.

It follows that the quasi-conserved quantities quantities that
characterize the orbit of the LG, i.e. $e$, $l$ and $\lambda$, do not
challenge the standard $\Lambda$CDM model, but that model is in
tension with regard to the particular phase on the orbit of the LG,
namely the actual values of the radial and tangential velocities. This
might hint to a problem of the $\Lambda$ CDM model to reproduce the
observed LG.

@@

\begin{table}
\caption{Summary of results from $\Lambda$CDM.}
\begin{center}
\begin{tabular}{cccc}\hline
 &  & Full Sample & Reduced Sample\\
$v_{\rm M31,rad}$ &(\kms) & $-70\pm 10$ & $-90\pm 10$\\
$v_{\rm M31,tan}$ &(\kms) & $50\pm 10$ & $50\pm 10$\\
$\log_{10}\lambda$& & $-1.47\pm 0.13$& \\ 
\end{tabular}
\end{center}
\vspace{1mm}
Note: the observational uncertainties in the position vector
correspond to a $5\%$ in each component consistent with the
uncertainties in the distance \citep[see references
  in][]{vanderMarel08}

@@

\begin{table*}
\caption{Summary of the comparison against $\Lambda$CDM.}
\begin{center}
\begin{tabular}{ccccc}\hline
Physical & (\%) Pairs consistent & \% Pairs consistent & \% Pairs
consistent& (\%) Pairs consistent\\ 
& with observations & with observations & with LCDM & with LCDM\\ 
 & in full sample & in reduced sample & in full sample & in reduced sample\\ \hline
$v_{\rm r}$-$v_{\rm t}$ & (0.4\%) 8/1923 & (0\%)0/158 & (1\%)23/1923 & (8\%)13/158\\
$e_{\rm tot}$-$l_{\rm orb}$ & (15\%)298/1923 & (24\%)38/158 & - & -\\
$\log_{10}\lambda$ & (13\%)257/1923 & (9\%)15/158 & - & -\\
$r_{\rm t}=v_{\rm t}/v_{\rm r}$& (12\%)242/1923 & (8\%)13/158 & -& -\\
\end{tabular}
\end{center}
\vspace{1mm}
Note: the observational uncertainties in the position vector
correspond to a $5\%$ in each component consistent with the
uncertainties in the distance \citep[see references
  in][]{vanderMarel08}

\label{table:2}
\end{table*}

@@

Figure \ref{fig:rt} summarizes the central finding of this
Letter. Most of the LG-like pairs (left panel) in the Bolshoi
Simulation and the pairs in the constrained simulation have radial and
tangential velocities notably different from the observational
constraints.  The most probable velocities in the simulation are
around $v_{\rm rad,\Lambda CDM} = -70\pm5$\kms and $v_{\rm tan,\Lambda
  CDM} = 50\pm5$ \kms, where the uncertainties in these values reflect
the minimum grid size needed to obtain robust statistics from the 2D
histogram. There are $23$ pairs out of the full LG-like pairs sample
centered in those values within the same range of absolute
uncertainties allowed by observations (i.e. $\sigma_{\rm tan}=17$\kms
and $\sigma_{\rm rad}=4$\kms.) In contrast, there are only $8$ pairs
in the interval allowed by observations.

The left panel in the same Figure \ref{fig:rt} shows the results for a
more restrictive selection of objects in the LG-like sample, reducing
the number of objects from  $1923$   to $158$ and their most probable
configuration is $v_{\rm rad,\Lambda CDM} = -50\pm5$\kms and $v_{\rm
  tan,\Lambda CDM} = 50\pm5$\kms with $5$ objects within that interval
and none consistent with observations. In this case none of the pairs
are found in the interval allowed by observations. The high eccentric
orbit of the observed LG  constitutes an unlikely configurations for
the $\Lambda$CDM LG-like objects. This holds for the full and the
subsample of objects.

From Figure \ref{fig:rt} it is also clear that there is a significant
number of pairs with a high tangential-to-radial velocity ratio. The
peak in the pair number density is located around a region of $f_{\rm
  t}\equiv v_{\rm tan \Lambda CDM}/v_{\rm rad \Lambda CDM}\sim 0.7$,
while the observations suggest $f_{\rm t}<0.32$. In the Bolshoi
Simulation, out of $1923$ LG-like pairs we find that $242$ pairs
($\sim 12\%$ of the sample) have values for $f_{\rm t}$ consistent
with that constraint, for the more restrictive sample there are $13$
pairs within that constraint ($\sim 8\%$ of the sample). The three
pairs from the constrained realizations are also outside the region of
the most probable values in a $\Lambda$CDM cosmology with
tangential-to-radial ratios of $f_{\rm t}= 0.35, 0.45, 0.73$.

The difference between the left and right panels of Figure
\ref{fig:rt} reflects  the tension found from the full and the sub-
sample of the LG-like objects. {\bf Yehuda, What tension??? The
  tension comes between theory and observation, not between the two
  theoretical results.}

\begin{figure*}
\begin{center}
\includegraphics[keepaspectratio=true,width=0.46\textwidth]{./figures/test_rt_BDM.pdf}
\includegraphics[keepaspectratio=true,width=0.46\textwidth]{./figures/test_rt_BDM_narrow.pdf}
\caption{Histograms of the radial and tangential velocities for
  LG-like halo pairs in the Bolshoi simulation. The left panel
  corresponds to the general LG-like pair simple, the right panel
  corresponds to the subsample with cuts on pair separation and total
  mass, matching the observational constraints. The squared panel
  shows as a shaded histogram the number of pairs on the radial and
  tangential velocities plane. The rectangular panels show the
  histograms when only one of the velocity components is used to bin
  the data. The half ellipse corresponds to the observed observational
  constraints for M31 in the Galactocentric rest frame: $v_{\rm
    rad}=109.2\pm 4.4$ km s$^{11}$ and $v_{\rm tan}< 34.4$ km s$^{-1}$
  \citep{vanderMarel12}. The circles represent the positions of the
  pairs from the constrained simulations. The highest number density
  in the left panel is obtained around $v_{\rm rad,\Lambda CDM} =
  -70\pm5$ \kms, $v_{\rm tan,\Lambda CDM} = 50\pm5$ \kms, for the
  right panel it is located around $v_{\rm rad,\Lambda CDM} = -50\pm5$
  \kms, $v_{\rm tan,\Lambda CDM} = 50\pm5$ } 
\label{fig:rt}
\end{center}

@@

@@

Conversely, there is an observational disadvantage. These quantities
are derived from very different kinds of observations, increasing the
uncertainties in their final determination.

A possible conflict between the likelihood of the LG in the
$\Lambda$CDM model as viewed from the radial and tangential velocities
as opposed to the angular momentum and energy perspectives would point
to a particular phase on the LG two-body orbit. Given the masses,
velocity and position of the observed MW and M31 (TAble \ref{table:1}
) the reduced energy and angular momentum are calculated. The Monte
Carlo sampling is used to to estimate the uncertainty in these
quantities.

Figure \ref{fig:EJ} presents the reduced  energy and angular momentum
distribution of the constrained and unconstrained LG-like  objects of
the full and subsample, following the   plotting conventions from
Figure \ref{fig:rt}.  The  1-$\sigma$ contour of Figure \ref{fig:EJ}
does not cross the zero level, as can be naively expected from Figure
\ref{fig:rt}. This does happen for the $\approx 80\%$ level
contour. The small difference between the two presentations of the
observational uncertainties stems from the Monte Carol sampling.

Inspection of Figure \ref{fig:EJ}  shows that the $e_{\rm tot}$  and
$l_{\rm orb}$ constraints are less restrictive compared with the
radial and tangential velocity constraints. For the unconstrained full
(sub-) sample 15\% (24\%)  of the LG-like pairs obey the 1-$\sigma$
observational constraints. All the constrained LGs of the full
samples are consistent with the 1-$sigma$ observational uncertainty.

However a reduction by a factor of $2$ in the uncertainty of the tangential velocity showing that the tangential velocity is below $17$\kms would 
bring these pairs and the $\Lambda$CDM expectation outside the
1-$\sigma$ observational uncertainty.

\begin{figure*}
\begin{center}
\includegraphics[keepaspectratio=true,width=0.46\textwidth]{./figures/test_lambda_BDM.pdf}
\includegraphics[keepaspectratio=true,width=0.46\textwidth]{./figures/test_lambda_BDM_narrow.pdf}
\caption{{\rm \label{fig:lambda} Normalized histograms of Peeble's
    spin parameter $\lambda$ for the pairs in the Bolshoi simulation
    and its infered values for the LG from the observational
    constraints from a Monte Carlo simulation. The left panel includes
    all the LG-pairs while the right panel includes a reduced sample
    of pairs that match the separation and total mass infered from
    Local Group observations. The vertical lines with the white dots
    represent the values inferred for pairs in the constrained
    simulations.}} 
\label{fig:lambda}
\end{center}
\end{figure*}

@@

Approximating the LG as two point masses we express the above
mentioned results in terms of the orbital angular momentum $l_{\rm
  orb}$ and the mechanical energy $e_{\rm tot}$ per unit of reduced
mass. We find that the uncertainties in the tangential velocity, the
square of the norm of the velocity and the total mass in the LG
produce poorer constraints on the number of simulated pairs that are
consistent with the observations. Nevertheless, in the case of the
LG-pair sample that also fulfills the separation and total mass
criteria there is a slight tension between simulation and
observation. A reduction by a factor of $2$ in the observational
uncertainty on the radial velocity would clarify this issue.

@@

Under the approximation of consevartion of the orbital angular
momentum and mechanical energy, we see that an agreement in the total
angular momentum and energy and a marked difference with the precise
balance between today's radial and tangential velocities could only be
explained if the initial conditions for the formation of the Local
Group are special in comparison to the initial conditions of any other
pair of dark matter halos in the $\Lambda$CDM cosmology.

We consider that reducing the uncertainties in the observations
relevant to this study are needed. Likewise, additional theoretical
work to understand the origin and implications of this result deserve
further study.

\label{sec:conclusions}
\section*{Acknowledgments}  
J.E.F-R acknowledges financial support from the Vicerrector\'{\i}a de
Investigaciones at Universidad de los Andes through its {\it Fondo de
  Apoyo a Profesores Asistentes} and the Peter and Patricia Gruber
Foundation through its fellowship administered by the International
Astronomical Union. J.E.F-R also acknowledges early discussions with
Alejandro Garc\'ia that motivated this work.

>>>>

\begin{abstract}
Recent observations constrained the tangential velocity of M31 with
respect to the Milky Way (MW) to be $v_{\rm M31,tan}<34.4$ \kms and
the radial velocity to be in the range $v_{\rm M31,rad}=-109\pm
4.4$\kms \citep{vanderMarel12}. In this study we use a large volume
high resolution N-body cosmologicalsimulation (BOLSHOI) to
statistically study this kinematics in the context of the $\Lambda$CDM
cosmology and three constrained simulations designed to reproduce the
Local Group (LG) and the local cosmic neighborhood. A LG is defined
here as  composed by two dominant separate halos hosting the MW and
the M31 galaxy fulfilling certain isolation criteria. An ensemble  of
all BOLSHOI pairs fulfilling this selection is constructed, and also a
subsample of it which obeys  more stringent observational constraints
on the mass and separation of of the MW and M31. The comparison of the
ensembles of simulated pairs with the observed LG has been done with
respect to the observed radial and tangential,  the reduced orbital
energy ($e$) and angular momentum ($l$) of the LG, and with respect to
the dimensionless spin parameter, $\lambda$. Our main results are: (i)
the prefered radial and tangential velocities for pairs in $\Lambda$CDM
are $v_{\rm r}=-80\pm 20$, $v_{\rm t}=50\pm 10$, (ii) pairs around
that region are $3$ to $13$ times more common than pairs within the
observational values, (iii) $15\%$ to $24\%$ of pairs in $\Lambda$CDM have
energy and angular momentum consistent with observations while (iv) $9\%$
to $13\%$ of pairs in the same sample show similar values to the inferred dimensionless spin
parameter. It follows that the quasi-conserved quantities quantities that
characterize the orbit of the LG, i.e. $e$, $l$ and $\lambda$, do not
challenge the standard $\Lambda$CDM model, but that model is in
tension with regard to the particular phase on the orbit of the LG,
namely the actual values of the radial and tangential velocities. This
might hint to a problem of the $\Lambda$CDM model to reproduce the
observed LG.

@@

\begin{table}
\caption{Summary of results from $\Lambda$CDM.}
\begin{center}
\begin{tabular}{cccc}\hline
 &  & Full Sample & Reduced Sample\\
$v_{\rm M31,rad}$ &(\kms) & $-70\pm 10$ & $-90\pm 10$\\
$v_{\rm M31,tan}$ &(\kms) & $50\pm 10$ & $50\pm 10$\\
$\log_{10}\lambda$& & $-1.47\pm 0.13$& $-1.34\pm 0.12$\\\hline
\end{tabular}
\end{center}
\vspace{1mm}
Note: the observational uncertainties in the position vector
correspond to a $5\%$ in each component consistent with the
uncertainties in the distance \citep[see references
  in][]{vanderMarel08}

@@

@@

Figure \ref{fig:rt} summarizes the central finding of this
Letter. Most of the pairs in the two samples constructed from the
Bolshoi Simulation have radial and tangential velocities notably
different from the observational constraints.

The most probable radial and tangential velocities in $\Lambda$CDM are
summarized in Table \ref{table:2}. For the full sample we have $v_{\rm
  rad,\Lambda CDM} = -70\pm10$\kms and $v_{\rm tan,\Lambda CDM} =
50\pm20$ \kms. For the reduced sample $v_{\rm
  rad,\Lambda CDM} = -90\pm10$\kms and $v_{\rm tan,\Lambda CDM} =
50\pm20$ \kms, where the uncertainties in these values reflect
the minimum grid size needed to obtain robust statistics for the 2D
histogram.

The number of pairs withing the range of uncertainty of observations
are listed in Table \ref{table:3} (columns 2 and 3). In the same table
(columns 4 and 5) we summarize the number of pairs around the
$\Lambda$CDM values within the same range of absolute observational
uncertainty (i.e. $\sigma_{\rm tan}=17$\kms and $\sigma_{\rm
  rad}=4$\kms.)

From these results we infer that the pairs around the preferred phase
space for $\Lambda$CDM are at least $13$ times more common than pairs
with the observed velocities for the Local group. We highlight that
this is a lower bound given that in the reduced sample none of the
pairs are found in the interval allowed by observations. The high
eccentric orbit of the observed LG  constitutes an unlikely
configurations for the $\Lambda$CDM LG-like objects. This holds for
the full and the subsample of objects.

From Figure \ref{fig:rt} it is also clear that there is a significant
number of pairs with a high tangential-to-radial velocity ratio. The
peak in the pair number density is located around a region of $f_{\rm
  t}\equiv v_{\rm tan \Lambda CDM}/v_{\rm rad \Lambda CDM}\sim 0.7$,
while the observations suggest $f_{\rm t}<0.32$. As summarized in Table
\ref{table:3} we find that only between $8\%$ to $12\%$ of the pairs are
consistent with the observational constraint. The three
pairs from the constrained realizations are also outside the region of
the most probable values in a $\Lambda$CDM cosmology with
tangential-to-radial ratios of $f_{\rm t}= 0.35, 0.45, 0.73$.

\begin{figure*}
\begin{center}
\includegraphics[keepaspectratio=true,width=0.46\textwidth]{./figures/test_rt_BDM.pdf}
\includegraphics[keepaspectratio=true,width=0.46\textwidth]{./figures/test_rt_BDM_narrow.pdf}
\caption{Histograms of the radial and tangential velocities for
  LG-like halo pairs in the Bolshoi simulation. The left (right) panel
  correspond to the full (reduced) pair sample.
The squared panel shows as a shaded histogram the number of pairs on
the radial and tangential velocities plane. The rectangular panels show the
  histograms when only one of the velocity components is used to bin
  the data. The half ellipse corresponds to the observed observational
  and the circles represent the positions of the
  pairs from the constrained simulations. The locii for the highest
  density regions are listed in Table \ref{table:2}.} 
\label{fig:rt}
\end{center}

@@

\begin{table*}
\caption{Summary of the comparison of the observational results
  against $\Lambda$CDM.}
\begin{center}
\begin{tabular}{ccccc}\hline
Physical & (\%) Pairs consistent & \% Pairs consistent & \% Pairs
consistent& (\%) Pairs consistent\\ 
property & with observations & with observations & with $\Lambda$CDM & with $\Lambda$CDM\\ 
 & in full sample & in reduced sample & in full sample & in reduced sample\\ \hline
$v_{\rm r}$-$v_{\rm t}$ & (0.4\%) 8/1923 & (0\%)0/158 & (1\%)23/1923 & (8\%)13/158\\
$e_{\rm tot}$-$l_{\rm orb}$ & (15\%)298/1923 & (24\%)38/158 & - & -\\
$\log_{10}\lambda$ & (13\%)257/1923 & (9\%)15/158 & - & -\\
$r_{\rm t}=v_{\rm t}/v_{\rm r}$& (12\%)242/1923 & (8\%)13/158 & -& -\\\hline
\end{tabular}
\end{center}
\vspace{1mm}
Note: In columns 4 and 5 the number of pairs around the prefered region
in $v_{\rm r}-v_{\rm t}$ in $\Lambda$CDM (as shown in
Fig. \ref{figure:rt} and summarized in Table \ref{table:2}) are
calculated within the same observational absolute uncertainty.
\label{table:3}
\end{table*}

@@

However, this formulation has an observational disadvantage. The
reduced energy and angular momentum are derived from very different
kinds of observations, increasing the uncertainties in their final
determination.

A possible conflict between the likelihood of the LG in the
$\Lambda$CDM model as viewed from the radial and tangential velocities
as opposed to the angular momentum and energy perspectives would point
to a particular phase on the LG two-body orbit. Given the masses,
velocity and position of the observed MW and M31 (Table \ref{table:1}
) the reduced energy and angular momentum are calculated. The Monte
Carlo sampling is used to to estimate the uncertainty in these
quantities.

Figure \ref{fig:EJ} presents the reduced  energy and angular momentum
distribution of the constrained and unconstrained LG-like  objects of
the full and subsample, following the   plotting conventions from
Figure \ref{fig:rt}.  The  1-$\sigma$ contour of Figure \ref{fig:EJ}
does not cross the zero level, as can be naively expected from Figure
\ref{fig:rt}. This happens for the $\approx 80\%$ level
contour. The small difference between the presentations of the
observational uncertainties in Figure \ref{fig:rt} versus Figure
\ref{fig:EJ} stems from the Monte Carlo sampling.

Visual inspection of Figure \ref{fig:EJ}  shows that the $e_{\rm tot}$  and
$l_{\rm orb}$ constraints are less restrictive compared with the
radial and tangential velocity constraints. As summarize in Table
\ref{table:3}, for the unconstrained full (sub-) sample 15\% (24\%)
of the LG-like pairs obey the 1-$\sigma$ observational
constraints.

All the constrained LGs of the full samples are
consistent with the 1-$\sigma$ observationaluncertainty.  However a
reduction by a factor of $2$ in the uncertainty of the tangential
velocity showing that the tangential velocity is below $17$\kms would
bring these pairs and the $\Lambda$CDM expectation outside the 1-$\sigma$ observational uncertainty.

\begin{figure*}
\begin{center}
\includegraphics[keepaspectratio=true,width=0.46\textwidth]{./figures/test_lambda_BDM.pdf}
\includegraphics[keepaspectratio=true,width=0.46\textwidth]{./figures/test_lambda_BDM_narrow.pdf}
\caption{{\rm \label{fig:lambda} Normalized histograms of Peeble's
    spin parameter $\lambda$ for the pairs in the Bolshoi simulation
    and its infered values for the LG from the observational
    constraints from a Monte Carlo simulation. The left (right) panel
    corresponds to the full (reduced) sample. The vertical lines with
    the white dots   represent the values inferred for pairs in the
    constrained 
    simulations.}} 
\label{fig:lambda}
\end{center}
\end{figure*}

@@

Approximating the LG as two point masses we express the above
mentioned results in terms of the orbital angular momentum $l_{\rm
  orb}$ and the mechanical energy $e_{\rm tot}$ per unit of reduced
mass. We find that the uncertainties in the tangential velocity, the
square of the norm of the velocity and the total mass in the LG
are less constraining on the number of simulated pairs that are
consistent with the observations. Nevertheless, in the case of the
LG-pair sample that also fulfills the separation and total mass
criteria there is a slight tension between simulation and
observation. A reduction by a factor of $2$ in the observational
uncertainty on the radial velocity would clarify this issue.

@@

Under the approximation of consevartion of the orbital angular
momentum and mechanical energy, we see that an agreement in the total
angular momentum and energy and a marked difference with the precise
balance between today's radial and tangential velocities could only be
explained if the initial conditions for the formation of the Local
Group are special in comparison to the initial conditions of any other
pair of dark matter halos in the $\Lambda$CDM cosmology. This opens a
new window into the question of how unique, if at all, is the LG in a
cosmological context. This will continue to be studies within the
CLUES framework.

\label{sec:conclusions}
\section*{Acknowledgments}  
JEF-R acknowledges financial support from the Vicerrector\'{\i}a de
Investigaciones at Universidad de los Andes through its {\it Fondo de
  Apoyo a Profesores Asistentes} and the Peter and Patricia Gruber
Foundation through its fellowship administered by the International
Astronomical Union. JEF-R also acknowledges early discussions with
Alejandro Garc\'ia that motivated this work. YH has been supported by
the ISF (1013/12)
